---
title: "Subsetting Datasets in R Tutorial"
author: "Tom Jeon"
date: "May 3, 2018"
output:
  html_document:
    self_contained: false
---

```{r, include=FALSE}
tutorial::go_interactive()
```


## Subsetting Datasets in R Tutorial

### Subsetting datasets is a crucial skill for any data professional. Learn and practice subsetting data in this quick interactive tutorial!

Whether you're comparing how different demographics respond to marketing campaigns, zooming in on a specific time period, or pulling information about a select few products of many, subsetting datasets is necessary when you want to take a closer look at your data. R is a great tool that makes subsetting data easy and intuitive. By the end of this tutorial, you'll have the know-how to extract the information you want from your dataset.

Most commonly, data are in rows and columns, so subsetting your data is simply selecting certain rows and/or columns based on some criteria you have in mind. In general, there are three ways to subset the rows and columns of your dataset—by index, by name, and often most usefully, by value.

### Subsetting rows and columns by index

One way to subset your rows and columns is by your dataset's indices. What does this mean? Well, in English, this is the same as describing your rows and columns as "the first row", "all rows in second and fifth columns", or "the first row in second to fifth columns". Let's see how you can specify such phrases using a dataset called `iris` in R:

```{r ex="create_a", type="sample-code"}
# "The first row":
iris[1, ]

# "All rows in second and fifth columns":
iris[, c(2, 5)]

# "The first row in second to fifth columns":
iris[1, 2:5]
```

As you can see, square brackets are used after your dataset object. The rows of your dataset is specified as the first element inside the square brackets and the columns of your dataset is specified as the second, separated by a comma:

```
data[rows, columns]
```

### Subsetting rows and columns by name

In R, the rows and columns of your dataset have name attributess. Row names are rarely used and by default provide indices—integers numbering from 1 to the number of rows of your dataset—just like what you saw in the previous section. In fact, if you called `rownames()` on the `iris` dataset, you will see that these are just indexed from 1 to 150:

```
> rownames(iris)
[1] "1"   "2"   "3"   "4"   "5"   "6"   "7"   "8"   "9"   "10"  "11"  "12"  "13"  "14" 
[15] "15"  "16"  "17"  "18"  "19"  "20"  "21"  "22"  "23"  "24"  "25"  "26"  "27"  "28" 
[29] "29"  "30"  "31"  "32"  "33"  "34"  "35"  "36"  "37"  "38"  "39"  "40"  "41"  "42" 
[43] "43"  "44"  "45"  "46"  "47"  "48"  "49"  "50"  "51"  "52"  "53"  "54"  "55"  "56" 
[57] "57"  "58"  "59"  "60"  "61"  "62"  "63"  "64"  "65"  "66"  "67"  "68"  "69"  "70" 
[71] "71"  "72"  "73"  "74"  "75"  "76"  "77"  "78"  "79"  "80"  "81"  "82"  "83"  "84" 
[85] "85"  "86"  "87"  "88"  "89"  "90"  "91"  "92"  "93"  "94"  "95"  "96"  "97"  "98" 
[99] "99"  "100" "101" "102" "103" "104" "105" "106" "107" "108" "109" "110" "111" "112"
[113] "113" "114" "115" "116" "117" "118" "119" "120" "121" "122" "123" "124" "125" "126"
[127] "127" "128" "129" "130" "131" "132" "133" "134" "135" "136" "137" "138" "139" "140"
[141] "141" "142" "143" "144" "145" "146" "147" "148" "149" "150"

> nrow(iris)
[1] 150
```

Row names are more common in smaller datasets that would make each observation in your dataset easily identifiable. For example, a doctor could have a small dataset containing health information of his or her patients, and the row names of this dataset might be the names of the patients.

Column names on the other hand, are ubiquitous to almost any dataset. You can access these with the `colnames()` function or the `names()` function:

```
> colnames(iris)
[1] "Sepal.Length" "Sepal.Width"  "Petal.Length" "Petal.Width"  "Species"     

> names(iris)
[1] "Sepal.Length" "Sepal.Width"  "Petal.Length" "Petal.Width"  "Species" 
```

To subset your dataset by the names of your rows and columnt, simply use the square brackets again prefixed by your dataset object:

```{r ex="create_b", type="sample-code"}
# Sepal width of the fifth observation
iris["5", "Sepal.Width"]

# Sepal width and petal width
iris[, c("Sepal.Width", "Petal.Width")]
```

It's important to note that both the row and column names are stored as characters, so using single or double quotes are absolutely necessary!

### Subsetting rows and columns by value

Subsetting your rows and columns by value often allows the most flexibility. For example, you can extract the data on the setosa iris using a conditional statement like this:

```
> iris[iris$Species == "setosa", ]
Sepal.Length Sepal.Width Petal.Length Petal.Width Species
1           5.1         3.5          1.4         0.2  setosa
2           4.9         3.0          1.4         0.2  setosa
3           4.7         3.2          1.3         0.2  setosa
4           4.6         3.1          1.5         0.2  setosa

...

47          5.1         3.8          1.6         0.2  setosa
48          4.6         3.2          1.4         0.2  setosa
49          5.3         3.7          1.5         0.2  setosa
50          5.0         3.3          1.4         0.2  setosa
```

Conditional statements like `iris$Species == "setosa"` belong in the row element in the square brackets (i.e. the first element before the comma). In addition to the conditional statement in the first element, you can specify columns by index or name in the second element. In the console below, try selecting just the sepal measurements of the setosa iris:

```{r ex="create_c", type="sample-code"}
# Print sepal measurements for setosa

```

```{r ex="create_c", type="solution"}
# Print sepal measurements for setosa (by name)
iris[iris$Species == "setosa", c("Sepal.Length", "Sepal.Width")]

# Print sepal measurements for setosa (by index)
iris[iris$Species == "setosa", 1:2]
```

```{r ex="create_c", type="sct"}
test_or(
  test_output_contains('iris[iris$Species == "setosa", c("Sepal.Length", "Sepal.Width")]',
                       incorrect_msg = "Select just the `Sepal.Length` and `Sepal.Width` columns for the setosa iris either by name or index."), 
  test_output_contains('iris[iris$Species == "setosa", c("Sepal.Width", "Sepal.Length")]',
                       incorrect_msg = "Select just the `Sepal.Length` and `Sepal.Width` columns for the setosa iris either by name or index."))
success_msg("Great work!")
```

### Practice exercises

Let's try practicing what you've learned thus far!

**Select the observations for which the sepal width is greater than the petal length.**

```{r ex="create_d", type="sample-code"}
# Which observations have greater sepal length than petal length?

```

```{r ex="create_d", type="solution"}
# Which observations have greater sepal width than petal length?
iris[iris$Sepal.Width > iris$Petal.Length, ]
```

```{r ex="create_d", type="sct"}
test_output_contains('iris[iris$Sepal.Width > iris$Petal.Length, ]',
                     incorrect_msg = "Not quite! Select the observations for which values for `Sepal.Width` is larger than those for `Petal.Length`.")
success_msg("Nice! Turns out this is the case only for the setosa iris.")
```

```{r ex="create_d", type="hint"}
Try using the `iris$Sepal.Width > iris$Petal.Length` conditional in the square brackets!
```

**Select the flowers that have bigger than average sepal width for the versicolor iris.**

```{r ex="create_e", type="sample-code"}
# Store the versicolor data as an object called versicolor


# Select the versicolor iris that have bigger than average sepal width

```

```{r ex="create_e", type="solution"}
# Store the versicolor data as an object called versicolor
versicolor <- iris[iris$Species == "versicolor", ]

# Select the versicolor iris that have bigger than average sepal width
versicolor[versicolor$Sepal.Width > mean(versicolor$Sepal.Width), ]
```

```{r ex="create_e", type="sct"}
test_object('versi', undefined_msg = "Did you define the `versicolor` object?", 
            incorrect_msg = "Select where `Species` is `\"versicolor\"`.")
test_output_contains('versicolor[versicolor$Sepal.Width > mean(versicolor$Sepal.Width), ], ]',
                     incorrect_msg = "Not quite! Try using the `mean()` function to find the average sepal width.")
success_msg("Great work!")
```

```{r ex="create_e", type="hint"}
Use the `mean()` function to find the average sepal width, then use a conditional to select only those with bigger values.
```

**Report the average sepal width for virginica irises that have a petal length smaller than 5 centimeters.**

```{r ex="create_f", type="sample-code"}
# You're free to approach this problem however you like!

```

```{r ex="create_f", type="solution"}
# Store the virginica data as an object called virginica
virginica <- iris[iris$Species == "virginica", ]

# Average sepal width for virginica irises that have a petal length smaller than 5
mean(virginica[virginica$Petal.Length < 5, ]$Sepal.Width)
```

```{r ex="create_f", type="sct"}
test_output_contains('2.8',
                     incorrect_msg = "Incorrect! This is a tough one!")
success_msg("Excellent!")
```

```{r ex="create_f", type="hint"}
First try subsetting for all virginica irises then for where `Petal.Length` is smaller than 5. You can find the average by calling the `mean()` function.
```

### Recap

In this tutorial, you:

- Learned about subsetting your data frame by index. Rows and columns are indexed as integers from 1 to the number of rows and columns, respectively.
- Learned about subsetting your data frame by name. You learned that row names are rarely specified and that column names are of character types.
- Learned to use conditional statements in the row element inside square brackets to subset your data frame by value.
- Learned to combine these methods to allow more flexible subsetting (e.g. using conditionals for rows and subsetting by index or name for columns).

Subsetting is just one way to manipulate your dataset. If you want to learn more about other ways to extract insights, such as grouping, summarizing, and merging, check out [these courses](https://www.datacamp.com/courses/tech:r/topic:data_manipulation) on DataCamp!